---
title: "Base Graphics"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
```

One of the greatest strengths of R, relative to other programming languages, is the ease with which we can create publication-quality graphics. In this lesson, you'll learn about base graphics in R.

We do not cover the more advanced portions of graphics in R in this lesson. For now, we're excluding `lattice`, `ggplot2` and `ggvis`. They will appear later, though. 

There is a school of thought that this approach is backwards, that we should teach `ggplot2` first. See http://varianceexplained.org/r/teach_ggplot2_to_beginners/ for an outline of this view.

Load the included data frame `cars` with `data(cars)`.

```{r a, exercise = TRUE}

```
```{r a-solution}
data(cars)
```

To fix ideas, we will work with simple data frames. Our main goal is to introduce various plotting functions and their arguments. All the output would look more interesting with larger, more complex data sets.


Pull up the help page for `cars`.

```{r prepare-b}
data(cars)
```
```{r b, exercise=TRUE}

```
```{r b-solution}
?cars
```

As you can see in the help page, the `cars` data set has only two variables: speed and stopping distance. Note that the data is from the 1920s. Run `head()` on `cars`.

```{r c, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r c-solution}
head(cars)
```

Before plotting, it is always a good idea to get a sense of the data. Key R commands for doing so include `dim()`, `names()`, `head()`, `tail()` and `summary()`.

Run `plot()` on `cars`.

```{r d, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r d-solution}
plot(cars)
```

As always, R tries very hard to give you something sensible given the information that you have provided to it. First, R notes that the data frame you have given it has just two columns, so it assumes that you want to plot one column versus the other.

Second, since we do not provide labels for either axis, R uses the names of the columns. Third, it creates axis tick marks at nice round numbers and labels them accordingly. Fourth, it uses the other defaults supplied in `plot()`.

We will now spend some time exploring `plot`, but many of the topics covered here will apply to most other R graphics functions. Note that 'plot' is short for scatterplot.

Look up the help page for `plot()`.

```{r e, exercise = TRUE}

```
```{r e-solution}
?plot
```

The help page for `plot()` highlights the different arguments that the function can take. The two most important are `x` and `y`, the variables that will be plotted. For the next set of questions, include the argument names in your answers. That is, do not type `plot(cars$speed, cars$dist)`, although that will work. Instead, use `plot(x = cars$speed, y = cars$dist)`.

Use `plot()` to show `speed` on the x-axis and `dist` on the y-axis from the `cars` data frame. Use the form of `plot` in which vectors are explicitly passed in as arguments for `x` and `y`.

```{r f, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r f-solution}
plot(x = cars$speed, y = cars$dist)
```

Note that this produces a slightly different answer than `plot(cars)`. In this case, R is not sure what you want to use as the labels on the axes, so it just uses the arguments which you pass in, data frame name and dollar signs included.

Note that there are other ways to call the plot command, i.e., using the "formula" interface. For example, we get a similar plot to the above with `plot(dist ~ speed, cars)`. However, we will wait till later in the lesson before using the formula interface.

Use `plot()` to show `dist` on the `x-axis` and speed on the `y-axis` from `cars`. This is the opposite of what we did above.

```{r g, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r g-solution}
plot(x = cars$dist, y = cars$speed)
```

It probably makes more sense for speed to go on the x-axis since stopping distance is a function of speed more than the other way around. So, for the rest of the questions in this portion of the lesson, always assign the arguments accordingly.

In fact, you can assume that the answers to the next few questions are all of the form `plot(x = cars$speed, y = cars$dist, ...)` but with various arguments used in place of the `...`.

Recreate the plot with the label of the x-axis set to "Speed".

```{r h, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r h-solution}
plot(x = cars$speed, y = cars$dist, xlab = "Speed")
```

Recreate the plot with, instead, the label of the y-axis set to "Stopping Distance".

```{r i, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r i-solution}
plot(x = cars$speed, y = cars$dist, ylab = "Stopping Distance")
```

Recreate the plot with both "Speed" and "Stopping Distance" as axis labels.

```{r j, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r j-solution}
plot(x = cars$speed, 
     y = cars$dist, 
     xlab = "Speed", 
     ylab = "Stopping Distance")
```

The reason that `plots(cars)` worked at the beginning of the lesson was that R was smart enough to know that the first element (i.e., the first column) in cars should be assigned to the x argument and the second element to the y argument. To save on typing, the next set of answers will all be of the form, `plot(cars, ...)` with various arguments added.

For each question, we will only want one additional argument at a time. Of course, you can pass in more than one argument when doing a real project.

Plot cars with only a main title of "My Plot". Note that the argument for the main title is "main" not "title".

```{r k, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r k-solution}
plot(cars, main = "My Plot")
```

Plot cars with only a sub title of "My Plot Subtitle".

```{r l, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r l-solution}
plot(cars, sub = "My Plot Subtitle")
```

The plot help page (`?plot`) only covers a small number of the many arguments that can be passed in to `plot()` and to other graphical functions. To begin to explore the many other options, look at `?par`. Let's look at some of the more commonly used ones. Continue using `plot(cars, ...)` as the base answer to these questions.

Plot `cars` so that the plotted points are colored red. (Use `col = 2` to achieve this effect.)

```{r m, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r m-solution}
plot(cars, col = 2)
```

Plot cars while limiting the x-axis to 10 through 15.  (Use `xlim = c(10, 15)` to achieve this effect.)

```{r n, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r n-solution}
plot(cars, xlim = c(10, 15))
```

You can also change the shape of the symbols in the plot. The help page for points (`?points`) provides the details.

Plot `cars` using triangles.  (Use `pch = 2` to achieve this effect.)

```{r o, exercise = TRUE, exercise.setup = "prepare-b"}

```
```{r o-solution}
plot(cars, pch = 2)
```

Arguments like `col` and `pch` may not seem very intuitive---because they aren't. So many people use more modern packages, like `ggplot2`, for creating their graphics in R.

It is, however, useful to have an introduction to base graphics because many of the idioms in `lattice` and `ggplot2` are modeled on them.

Let's now look at some other functions in base graphics that may be useful, starting with boxplots.

Load the `mtcars` data frame.

```{r prepare-p}
data(mtcars)
```
```{r p, exercise=TRUE}

```
```{r p-solution}
data(mtcars)
```

Anytime that you load up a new data frame, you should explore it before using it. Use any code box to play around however you like.

Look up the help page for `boxplot()`.

```{r q, exercise = TRUE}

```
```{r q-solution}
?boxplot
```

Instead of adding data columns directly as input arguments, as we did with `plot()`, it is often handy to pass in the entire data frame. This is what the "data" argument in `boxplot()` allows.

`boxplot()`, like many R functions, also takes a "formula" argument, generally an expression with a tilde ("~") which indicates the relationship between the input variables. This allows you to enter something like `mpg ~ cyl` to plot the relationship between `cyl` (number of cylinders) on the x-axis and `mpg` (miles per gallon) on the y-axis.

Use `boxplot()` with `formula = mpg ~ cyl` and `data = mtcars` to create a box plot.

```{r r, exercise = TRUE, exercise.setup = "prepare-p"}

```
```{r r-solution}
boxplot(formula = mpg ~ cyl, data = mtcars)
```

The plot shows that `mpg` is much lower for cars with more cylinders. Note that we can use the same set of arguments that we explored with `plot()` above to add axis labels, titles and so on.

When looking at a single variable, histograms are a useful tool. `hist()` is the associated R function. Like `plot()`, `hist()` is best used by just passing in a single vector.

Use `hist()` with the vector `mtcars$mpg` to create a histogram.

```{r s, exercise = TRUE, exercise.setup = "prepare-p"}

```
```{r s-solution}
hist(mtcars$mpg)
```

In this lesson, you learned how to work with base graphics in R. The best place to go from here is to study the `ggplot2` package. If you want to explore other elements of base graphics, then (http://www.ling.upenn.edu/~joseff/rstudy/week4.html) provides a useful overview.

