---
title: "Variables"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
```

This video is about 5 and a half minutes long and contains an explanation of what a variable is and why it might be useful (it is also colorful and somewhat fascinating to watch the narrator, Sal, write with his colorful pens on the black background). I'll ask a question or two about variables next, and the answers are in the video.

![Variables explained by Sal Khan](https://www.youtube.com/watch?v=3v9w79NhsfI)

```{r a, echo=FALSE}
question("What is the BEST definition of a random variable?",
         answer("a method for mapping the outcomes of random processes to numbers", correct = TRUE),
         answer("something you use in algebra in place of actual values"),
         answer("the best way to give you a headache"),
         answer("the opposite of a specific variable")
)
```

You probably won't be surprised to know that different people have different names for variable types. I'll list names for variable types that I have come across.

### Scalar a.k.a. Measurement a.k.a. Numeric a.k.a. Continuous a.k.a. Quantitative Variables

Scalar variables represent things you can measure. They have many names, but it all boils down to the fact that the numbers exist on a scale rather than being names representing different states (such as "There" vs. "Not there" or "Yes" vs. "No" or "Red" vs. "Brown" vs. "Blue" vs. "Green" vs... etc.).

```{r b, echo=FALSE}
question("Which of the following represents a scalar variable?",
         answer("the length of a newborn person", correct = TRUE),
         answer("a child's grade level in school"),
         answer("a species of vegetable"),
         answer("ethnicity")
)
```

### Categorical a.k.a. Nominal a.k.a. Discrete Variables

Categorical variables represent things that fit into categories. They are represented by the names of their types, such as a type of vegetable, a vacation destination, or a dead vs. a living strawberry.

```{r c, echo=FALSE}
question("Which of the following is NOT a categorical variable?",
         answer("gender"),
         answer("the road name in a house's address", correct = TRUE), 
         answer("a strawberry's status"),
         answer("test score as a percentage")
)
```

### Binary a.k.a. Boolean Variables

Binary variables are a special case of categorical variables. They have two, no more and no less, categories.

```{r d, echo=FALSE}
question("Select a binary variable from the list.",
         answer("The number of pages in The Life and Opinions of Tristram Shandy, Gentleman"),
         answer("Answer (y/n)", correct = TRUE),
         answer("your birthday"),
         answer("the length of time it takes to wake up in the morning")
)
```

### Miscellany

```{r e, echo=FALSE}
question("Which of the following is a scalar variable?",
         answer("days of the week", correct = TRUE),
         answer("probability of getting heads when flipping a coin"),
         answer("bone density")
)
```

```{r f, echo=FALSE}
question("",
         answer(""),
         answer("", correct = TRUE),
         answer("")
)
```

Some authors divide scalar variables into two types, continuous variables such as the length of an isopod's antenna, which in theory have an infinite number of possible values (think how many decimal places you can use: that's right--infinity), and discrete (or meristic) variables, which have only whole number values.

Discrete variables represent things you can count, such as the number of spines on an isopod's antenna, and know that you haven't skipped any values in between.

You don't count them by saying, "One, two, two and a half, three," for instance. But the number of spines is 

Fortunately for us and for all budding statisticians, statistical tests for continuous scalar variables work well on discrete scalar variables, so you usually don't need to worry about the difference between continuous and discrete scalar variables.

```{r g, echo=FALSE}
question("Let's review. Say you have two types of isopods, those with 32 spines and those with 33 spines. There are no other types of isopods. What kind of variable is number of spines?",
         answer("scalar"),
         answer("binary", correct = TRUE),
         answer("categorical")
)
```

If you have only two possible values for a discrete value, you might treat the variable as nominal or, better, binary.

```{r h, echo=FALSE}
question("What kind of variable has two possible answers?",
         answer("scalar"),
         answer("binary", correct = TRUE),
         answer("categorical")
)
```

```{r i, echo=FALSE}
question("What kind of variable has five (and only five) possible answers, each of which can be labeled?",
         answer("binary"),
         answer("categorical", correct = TRUE),
         answer("scalar")
)
```
```{r j, echo=FALSE}
question("What kind of variable can have an infinite number of possible answers?",
         answer("categorical"),
         answer("scalar", correct = TRUE),
         answer("binary")
)
```
```{r k, echo=FALSE}
question("If you group together people's ages into age groups, you are changing what kind of variable into what other kind of variable?",
         answer("binary to categorical"),
         answer("scalar to categorical", correct = TRUE),
         answer("categorical to binary")
)
```
```{r l, echo=FALSE}
question("Knowing what types of variables you have in your data set gives you what advantage?",
         answer("you know how to analyze them"),
         answer("all of these", correct = TRUE),
         answer("you know how to summarize them"),
         answer("you know how to plot them")
)
```

In this lesson, you have been introduced to the all-important practice of figuring out what kind of variable you might be examining. We have considered binary, scalar, and categorical variables. Scalar variables are often called "measurement" variables; sometimes people refer to categorical variables as "nominal" variables. Sometimes people refer to binary variables as "Boolean".

We have not yet considered date variables, but if you explore the [R Programming lessons](), you will learn how to use them.



