---
title: "vapply and tapply"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(data.table)
library(dplyr)
load("flags.RData")
flag_colors <- flags[, 11:17]
flag_shapes <- flags[, 19:23]
shape_mat <- sapply(flag_shapes, range)
unique_vals <- lapply(flags, unique)
knitr::opts_chunk$set(echo = FALSE)
```


## More Loop Functions in R: `vapply()` and `tapply()`

In this lesson, you'll learn how to use `vapply()` and `tapply()`, each of which serves a very specific purpose within the Split-Apply-Combine methodology mentioned in [the previous lesson](/lapply-and-sapply/lapply-and-sapply. For consistency, we'll use the same dataset we used in the '`lapply` and `sapply`' lesson.

The Flags dataset from the UCI Machine Learning Repository contains details of various nations and their flags. More information may be found at http://archive.ics.uci.edu/ml/datasets/Flags.

I've stored the data in a variable called flags. If it's been a while since you completed the '`lapply` and `sapply`' lesson, you may want to reacquaint yourself with the data by using functions like `dim()`, `head()`, `str()`, and `summary()` at any prompt below. You can also type `viewinfo()` at the prompt to bring up some documentation for the `flags` dataset.

As you saw in the last lesson, `unique()` returns a vector of the unique values contained in the object passed to it. Therefore, `sapply(flags, unique)` returns a list containing one vector of unique values for each column of the `flags` dataset. Try it again now.

```{r a, exercise=TRUE}

```
```{r a-solution}
sapply(flags, unique)
```

What if you had forgotten how `unique()` works and mistakenly thought it returns the *number* of unique values contained in the object passed to it? Then you might have incorrectly expected `sapply(flags, unique)` to return a numeric vector, since each element of the list returned would contain a single number and `sapply()` could then simplify the result to a vector.

When working interactively (at the `>` prompt), this is not much of a problem, since you see the result immediately and will quickly recognize your mistake. However, when working non-interactively (e.g. writing your own functions), a misunderstanding may go undetected and cause incorrect results later on. Therefore, you may wish to be more careful and that's where `vapply()` is useful.

Whereas `sapply()` tries to 'guess' the correct format of the result, `vapply()` allows you to specify it explicitly. If the result doesn't match the format you specify, `vapply()` will throw an error, causing the operation to stop. This can prevent significant problems in your code that might be caused by getting unexpected return values from `sapply()`.

Try `vapply(flags, unique, numeric(1))`, which says that you expect each element of the result to be a numeric vector of length 1. Since this is NOT actually the case, *YOU WILL GET AN ERROR*.

```{r b, exercise = TRUE}

```
```{r b-solution}
vapply(flags, unique, numeric(1))
```

Recall from the previous lesson that `sapply(flags, class)` will return a character vector containing the class of each column in the dataset. Try that again now to see the result.

To open a more complete description of the dataset in a separate text file, type `viewinfo()`.

```{r c, exercise = TRUE}

```
```{r c-solution}
sapply(flags, class)
```

If we wish to be explicit about the format of the result we expect, we can use `vapply(flags, class, character(1))`. The `character(1)` argument tells R that we expect the class function to return a character vector of length 1 when applied to EACH column of the flags dataset. Try it now.

```{r d, exercise = TRUE}

```
```{r d-solution}
vapply(flags, class, character(1))
```

Note that since our expectation was correct (i.e. `character(1)`), the `vapply()` result is identical to the `sapply()` result---a character vector of column classes.

You might think of `vapply()` as being 'safer' than `sapply()`, since it requires you to specify the format of the output in advance, instead of just allowing R to 'guess' what you wanted. In addition, `vapply()` may perform faster than `sapply()` for large datasets. However, when doing data analysis interactively (at the prompt), `sapply()` saves you some typing and will often be good enough.

As a data analyst, you'll often wish to split your data up into groups based on the value of some variable, then apply a function to the members of each group. The next function we'll look at, `tapply()`, does exactly that.

Use `?tapply` to pull up the documentation.

```{r f, exercise = TRUE}

```
```{r f-solution}
?tapply
```

The `landmass` variable in our dataset takes on integer values between 1 and 6, each of which represents a different part of the world. Use `table(flags$landmass)` to see how many flags/countries fall into each group.

```{r g, exercise = TRUE}

```
```{r g-solution}
table(flags$landmass)
```

The `animate` variable in our dataset takes the value 1 if a country's flag contains an animate image (e.g. an eagle, a tree, a human hand) and 0 otherwise. Use `table(flags$animate)` to see how many flags contain an animate image.

```{r h, exercise = TRUE}

```
```{r h-solution}
table(flags$animate)
```

This tells us that 39 flags contain an animate object (`animate` = 1) and 155 do not (`animate` = 0).

If you take the arithmetic mean of a bunch of 0s and 1s, you get the proportion of 1s. Use `tapply(flags$animate, flags$landmass, mean)` to apply the mean function to `animate` separately for each of the six landmass groups, thus giving us the proportion of flags containing an animate image WITHIN each landmass group.

```{r i, exercise = TRUE}

```
```{r i-solution}
tapply(flags$animate, flags$landmass, mean)
```

The first landmass group (`landmass` = 1) corresponds to North America and contains the highest proportion of flags with an animate image (0.4194).

Similarly, we can look at a summary of population values (in round millions) for countries with and without the color red on their flag with `tapply(flags$population, flags$red, summary)`.

```{r j, exercise = TRUE}

```
```{r j-solution}
tapply(flags$population, flags$red, summary)
```

```{r aa, echo=FALSE}
question("What is the median population (in millions) for countries *without* the color red on their flag?",
      answer("27.6"),
      answer("3.0", correct = TRUE),
      answer("0.0"),
      answer("22.1"),
      answer("4.0"),
      answer("9.0")
)
```

Lastly, use the same approach to look at a summary of population values for each of the six landmasses.


```{r k, exercise = TRUE}

```
```{r k-solution}
tapply(flags$population, flags$landmass, summary)
```

Use the `head()` function to look at the first 6 lines of `flag_colors`.

```{r l, echo=FALSE}
question("What is the maximum population (in millions) for the fourth landmass group (Africa)?",
      answer("56.00", correct = TRUE),
      answer("119.0"),
      answer("5.0"),
      answer("157.00"),
      answer("1010.0")
)
```

In this lesson, you learned how to use `vapply()` as a safer alternative to `sapply()`, which is most helpful when writing your own functions. You also learned how to use `tapply()` to split your data into groups based on the value of some variable, then apply a function to each group. These functions will come in handy on your quest to become a better data analyst.