---
title: "Missing Values"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.completion = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

# Missing Values

Missing values play an important role in statistics and data analysis. Often, missing values must not be ignored, but rather they should carefully be studied to see if ]an underlying pattern exists, or some cause for their missingness that provides significant information.

In R, `NA` is used to represent any value that is ‘not available’ or ‘missing’ (in the statistical sense). In this lesson, we’ll explore missing values further.

Any operation involving `NA` generally yields `NA` as the result. To illustrate, let’s create a vector `c(44, NA, 5, NA)` and assign it to a variable `x`.

```{r a, exercise=TRUE}

```
```{r a-solution}
x <- c(44, NA, 5, NA)
```

Now, let’s multiply `x` by 3.

```{r prepare-b}
x <- c(44, NA, 5, NA)
```
```{r b, exercise=TRUE}

```
```{r b-solution}
x * 3
```

Notice that the elements of the resulting vector that correspond with the 'NA' values in x are also `NA`.

To make things a little more interesting, lets create a vector containing 1000 draws from a standard normal distribution with `y <- rnorm(1000)`.

```{r prepare-c}
set.seed(10)
```
```{r c, exercise=TRUE}

```
```{r c-solution}
y <- rnorm(1000)
```

Next, let’s create a vector containing 1000 'NA's with `z <- rep(NA, 1000)`.

```{r prepare-d}
set.seed(10)
y <- rnorm(1000)
set.seed(8)
```
```{r d, exercise=TRUE}

```
```{r d-solution}
z <- rep(NA, 1000)
```

Finally, let’s select 100 elements at random from these 2000 values (combining y and z) such that we don’t know how many 'NA's we’ll wind up with or what positions they’ll occupy in our final vector – `my_data <- sample(c(y, z), 100)`.

```{r prepare-e}
set.seed(10)
y <- rnorm(1000)
set.seed(8)
z <- rep(NA, 1000)
```
```{r e, exercise=TRUE}

```
```{r e-solution}
my_data <- sample(c(y, z), 100)
```

Let’s first ask the question of where our `NA`s are located in our data. The `is.na()` function tells us whether each element of a vector is `NA`. Call `is.na()` on `my_data` and assign the result to `my_na`. Then, print `my_na` to see what you came up with.

```{r prepare-f}
set.seed(10)
y <- rnorm(1000)
set.seed(8)
z <- rep(NA, 1000)
my_data <- sample(c(y, z), 100)
```
```{r f, exercise=TRUE}

```
```{r f-solution}
my_na <- is.na(my_data)
my_na
```

Everywhere you see a `TRUE`, you know the corresponding element of `my_data` is `NA`. Likewise, everywhere you see a `FALSE`, you know the corresponding element of `my_data` is one of our random draws from the standard normal distribution.

In our previous discussion of logical operators, we introduced the == operator as a method of testing for equality between two objects. So, you might think the expression `my_data == NA` yields the same results as `is.na()`. Give it a try.

```{r prepare-g}
set.seed(10)
y <- rnorm(1000)
set.seed(8)
z <- rep(NA, 1000)
my_data <- sample(c(y, z), 100)
my_na <- is.na(my_data)
```
```{r g, exercise=TRUE}

```
```{r g-solution}
my_data == NA
```

The reason you got a vector of all 'NA's is that 'NA' is not really a value, but just a placeholder for a quantity that is not available. Therefore the logical expression is incomplete and R has no choice but to return a vector of the same length as my_data that contains all NAs.

Don’t worry if that’s a little confusing. The key takeaway is to be cautious when using logical expressions anytime 'NA's might creep in, since a single 'NA' value can derail the entire thing.

So, back to the task at hand. Now that we have a vector, `my_na`, that has a `TRUE` for every `NA` and `FALSE` for every numeric value, we can compute the total number of 'NA's in our data.

The trick is to recognize that underneath the surface, R represents TRUE as the number 1 and FALSE as the number 0. Therefore, if we take the sum of a bunch of TRUEs and FALSEs, we get the total number of TRUEs.

Let’s give that a try here. Call the `sum()` function on `my_na` to count the total number of TRUEs in `my_na`, and thus the total number of 'NA's in `my_data`. Don’t assign the result to a new variable.

```{r h, exercise=TRUE, exercise.setup = "prepare-g"}

```
```{r h-solution}
sum(my_na)
```

Pretty cool, huh? Finally, let’s take a look at the data to convince ourselves that everything ‘adds up’. Print my_data to the console.

```{r i, exercise=TRUE, exercise.setup = "prepare-g"}

```
```{r i-solution}
my_data
```

Now that we’ve got 'NA's down pat, let’s look at a second type of missing value – `NaN`, which stands for ‘not a number’. To generate NaN, try dividing (using a forward slash) 0 by 0 now.

```{r j, exercise=TRUE}

```
```{r j-solution}
0/0
```

Let’s do one more, just for fun. In R, `Inf` stands for infinity. What happens if you subtract `Inf` from `Inf`?

```{r k, exercise=TRUE}

```
```{r k-solution}
Inf - Inf
```
