---
title: "Introduction to Bayes's Theorem"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
#' Assign some global variables 
prevalence <- 1 / 2500
sensitivity <- .962
specificity <- .998
cf_neg <- 1 - prevalence
false_pos_rate <- 1 - specificity
false_pos <- false_pos_rate * cf_neg
true_pos <- prevalence * sensitivity
all_pos <- true_pos + false_pos
conditional_probability <- (sensitivity * prevalence) / all_pos
```



The following example shows why a Bayesian approach to probability might be interesting to a medical researcher.

Let's look at a common probability question in medicine. Let's say that a child is discovered to have elevated immunoreactive tryspinogen, or IRT, a marker for cystic fibrosis, in blood smears hospital staff acquired on the first day of her life.

What is the probability that she actually has cystic fibrosis given that she tested positive for it? For the purpose of this analysis, we will not consider possible interaction variables or other predictors besides the positive test result.

First, set up some variables. 

### Prevalence

We know that 1 out of every 2500 live births involves a baby with cystic fibrosis. That's prevalence. Assign 1/2500 to the variable `prevalence`.

```{r a, exercise=TRUE}

```
```{r a-solution}
prevalence <- 1 / 2500
```

### Specificity

No matter how good a screening test is, there is a chance that a child with cystic fibrosis is not identified. Results that incorrectly classify someone as not having a disease are "false negatives." The chance of truly identifying all babies without cystic fibrosis is 99.8%. This is specificity, or the true negative rate. Assign the correct value to the variable `specificity`.

```{r b, exercise = TRUE}

```
```{r b-solution}
specificity <- .998
```

### Sensitivity

The test's sensitivity is the rate at which it correctly identifies those who have the disease. Another term for this is "probability of detection." The percentage of tested babies who are correctly classified as having the disease is 96.2%. Assign the correct number to the variable `sensitivity`.

```{r c, exercise = TRUE}

```
```{r c-solution}
sensitivity <- .962
```

### Prevalence

Our question was this: 

```{r d, echo=FALSE}
question("Given the fact that a baby tested positive for elevated RTF, what is the probability that she actually has cystic fibrosis?",
         answer("The same as sensitivity, which is .962"),
         answer(".038"),
         answer("1"),
         answer("These answers are all wrong (except this one)", correct = TRUE)
)
```
```{r e, echo=FALSE}
question("What variable was missing in all of the possible answers in the previous question?",
         answer("prevalence", correct = TRUE),
         answer("I have no idea"),
         answer("the fact that you're tired")
)
```

Why does prevalence matter? Let's break it down.

### Bayes's Theorem

There is a famous formula for answering our main question, the likelihood of having cystic fibrosis (or any disease) given elevated IRT (or any positive result to a screening test for a disease) called the Bayes's Theorem.

In sentence form, 

>Bayes's Theorem is the likelihood that a condition will occur given the observed data. 

In this case, Bayes's Theorem helps you figure out the likelihood that a child has cystic fibrosis given elevated IRT.

In formulaic terms, Bayes's Theorem is

>`P(A|B) = (P(B|A) * P(A)) / P(B)`.

The vertical lines in this formula mean "given." When you read the formula, therefore, you should say, "The probability of `A` given `B` is the probability of `B` given `A` times the probability of `A`, all over the probability of `B`."

The probability of `A` given `B` is the `sensitivity` of the test. The probability of `A` is the `prevalence`. The probability of `B` is the probability of getting a positive test result---regardless of whether or not one has the disease.

Let's map these variables to the concepts we have been discussion. 

```{r f, echo=FALSE}
question("What is P(A)",
      answer("the unconditional probability of having the disease, or prevalence", correct = TRUE),
      answer("the likelihood of having a positive test result"),
      answer("the likelihood of not having a disease given a negative test result")
)
```
```{r g, echo=FALSE}
question("Getting back to what we really want to know, what is `P(A|B)`?",
      answer("the likelihood of having a positive test result"),
      answer("the conditional probability of having a disease given a positive test result", correct = TRUE),
      answer("the likelihood of not having a disease given a negative test result")
)
```
```{r h, echo=FALSE}
question("What is `P(B)`?",
      answer("the probability of a positive test", correct = TRUE),
      answer("the probability of not having the disease"),
      answer("the probability of having the disease"),
      answer("the false positive rate")
)
```
```{r i, echo=FALSE}
question("Have we figured out yet how many babies have positive tests for IRT? This would be `P(B)` in Bayes's Theorem.",
      answer("No", correct = TRUE),
      answer("Yes")
)
```

### Negatives, True and False

Let's find `P(B)` in our example, the probability of having elevated IRT, by adding up those with and those without cystic fibrosis who have elevated IRT. We have numbers already that we can use for that.

```{r j, echo=FALSE}
question("Let's look at those WITHOUT cystic fibrosis who have elevated IRT. What represents that value?",
      answer("the false positive rate", correct = TRUE),
      answer("specificity"),
      answer("prevalence"),
      answer("sensitivity")
)
```

#### False Negative Rate

In order to calculate the false positive rate, we're going to need a variable for the number of babies born without cystic fibrosis. That number would be 1 minus something. What is it? Assign it to the variable `cf_neg`. Put parenthesis around the entire expression so you can see the actual number when you execute the code.

```{r k, exercise = TRUE}

```
```{r k-solution}
(cf_neg <- 1 - prevalence)
```

We know that specificity is the rate at which the test correctly identifies those without the disease. So what is `1 - specificity`? Is that the false positive rate? 

In a general sense, yes, it is. But we want to know the false positives for THIS situation. How do we find that number?

Let's create a variable called `false_pos_rate` and give it the value I just told you about, `1 - specificity`. Put parentheses around the statement so you can see the actual value.

```{r l, exercise = TRUE}

```
```{r l-solution}
(false_pos_rate <- 1 - specificity)
```

#### False Positives

Now let's find the false positives for this situation. We stored the total number of people born without cystic fibrosis in the variable `cf_neg`. We know that OF those people, `false_pos_rate` (which we just calculated to be 0.2% ) will be identified incorrectly as having cystic fibrosis. 

>Hint: Whenever you see a sentence of the form FIND X OF Y, you may remember from high school, you multiply X and Y together to find the answer. 

Do that here. Find `false_pos_rate` of `cf_neg`, store it in the variable `false_pos` and have a look at it, all in one line.

```{r m, exercise = TRUE}

```
```{r m-solution}
(false_pos <- false_pos_rate * cf_neg)
```

### Positives, True and False

We just looked at those WITHOUT cystic fibrosis who have elevated IRT. Create a statement that defines those WITH cystic fibrosis (`prevalence`) who ALSO test positive for it (`sensitivity`), using the same general method you just used to find the variable `false_pos` above. Assign it to the variable `true_pos`. This formula gets us from the concept of sensitivity to an actual number of sick kids who are correctly identified. Put parentheses around your formula so you can see the results right away.

```{r n, exercise = TRUE}

```
```{r n-solution}
(true_pos <- prevalence * sensitivity)
```

We now have what we need to calculate the likelihood of testing positive. Do so, and assign the value to `all_pos`. Use parentheses so you can see the value.

```{r o, exercise = TRUE}

```
```{r o-solution}
(all_pos <- true_pos + false_pos)
```

Now we have all the values we need to fill out Bayes's Theorem and get our answer.

## Plug and Chug

We can now plug in all the elements of Bayes's Theorem and get the answer to our question.

* We have mapped P(B|A), the probability of testing positive given that the disease is present, to the variable sensitivity.    
* We have mapped P(A), the probability of having cystic fibrosis, to the variable prevalence.    
* We have mapped P(B), the odds of testing positive whether or not the disease is present, to all_pos.    

>Remember that Bayes's Theorem is P(A|B) = (P(B|A) * P(A)) / P(B).

Now find the answer. How likely is a newborn with elevated IRT to have cystic fibrosis? Assign the answer to `conditional_probability`.

```{r p, exercise = TRUE}

```
```{r p-solution}
(conditional_probability <- (sensitivity * prevalence) / all_pos)
```

Call percentage() and pass it one argument, the variable `conditional_probability`.

```{r q, exercise = TRUE}

```
```{r q-solution}
percentage(conditional_probability)
```

```{r r, echo=FALSE}
question("So elevated IRT levels tell you what, exactly, about the child in your care?",
      answer("that she has a 96% chance of having cystic fibrosis"),
      answer("that she has a 99% chance of having cystic fibrosis"),
      answer("that she has an approximately 16% chance of having cystic fibrosis", correct = TRUE)
)
```
```{r s, echo=FALSE}
question("What does this mean?",
      answer("The odds of the baby having the disease are much lower than the sensitivity of the test", correct = TRUE),
      answer("The test is no good"),
      answer("The odds of the baby having the disease are the same as the sensitivity of the test"),
      answer("The odds of the baby having the disease are much higher than the specificity of the test")
)
```

If you have been paying very careful attention, you may have noticed that we can create a shorter formula to arrive at `conditional_probability`. Let's make sure. Type `identical((sensitivity * prevalence), true_pos)`.

```{r t, exercise = TRUE}

```
```{r t-solution}
identical((sensitivity * prevalence), true_pos)
```

So Bayes's theorem can actually be expressed very simply.

```{r label, echo=FALSE}
question(" How can Bayes's Theorem be expressed more simply than the formula we started out with?",
      answer("true_pos / all_pos", correct = TRUE),
      answer("all_pos / true_pos"),
      answer("sensitivity / all_pos")
)
```

The fact that the conditional probability of having cystic fibrosis is surprisingly low does not make the test a bad one. A jump from a .004% chance of having the disease to a 16% chance is a big jump and creates the need for a new plan of care, including the gold standard test for cystic fibrosis to establish a diagnosis. But the difference between 16% and 96% or even 99% likelihood of having cystic fibrosis might be a very important one for everyone who cares about the child during a harrowing time of uncertainty.

The important thing to remember is that there is a difference between the true positive RATE, which is general, and the true positive likelihood FOR THIS PROBLEM, which relates specifically to the prevalence. At the heart of Bayes's theorem for calculating the conditional probability of having a disease given a positive result lies prevalence. It alters everything. A rare disease will have a low conditional probability even if the screening test has high sensitivity and specificity. A common disease will have a high conditional probability even if the screening test is slightly less sensitive and specific.

That's the end of this introduction to conditional probability using Bayesian analysis. We used Bayes's theorem to interpret the meaning of testing positive for a relatively rare illness and found out that, since prevalence is such a small number, testing positive on a screening test isn't as scary as one might think. One should always wait for a gold standard test to validate the findings of a screening test that has a false positive rate that is higher than 0.

For more fun, look for more equivalences among the variables you created for this lesson. There is at least one more. Think about why it might be an equivalence---by chance, or will these two numbers always be equivalent? Will this make it easier for you to calculate conditional probability in the future?

Email me at braunsb@email.chop.edu for a printable version of this lesson.