---
title: "Exploratory Graphs"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
setwd("/Users/braunsb/Documents/Arcus Learning Plan Generator/Lessons")
library(learnr)
pollution <- readRDS("pollution.RData")
ppm <- pollution$pm25
```

# Exploratory Graphs

(Slides for this and other Data Science courses may be found at github [https://github.com/DataScienceSpecialization/courses/](https://github.com/DataScienceSpecialization/courses/). If you care to use them, they must be downloaded as a zip file and viewed locally. This lesson corresponds to 04_ExploratoryAnalysis/exploratoryGraphs.)

In this lesson, we'll discuss why graphics are an important tool for data scientists and the special role that exploratory graphs play in the field.

```{r a, echo=FALSE}
question("Which of the following would NOT be a good reason to use graphics in data science?",
         answer("To suggest modeling strategies"),
         answer("To find a color that best matches the shirt you're wearing", correct = TRUE),
         answer("To find patterns in data")
)
```

So graphics give us some visual form of data, and since our brains are very good at seeing patterns, graphs give us a compact way to present data and find or display any pattern that may be present.

```{r b, echo=FALSE}
question("Which of the following cliches captures the essence of graphics?",
         answer("A rose by any other name smells as sweet"),
         answer("A picture is worth a thousand words", correct = TRUE),
         answer("The apple doesn't fall far from the tree")
)
```

Exploratory graphs serve mostly the same functions as graphs in general. They help us find patterns in data and understand its properties. They suggest modeling strategies and help to debug analyses. 

However, we DON'T use exploratory graphs to communicate results.

Instead, exploratory graphs are the _initial_ step in an investigation, the "quick and dirty" tool used to point the data scientist in a fruitful direction. A scientist might need to make a lot of exploratory graphs in order to develop a personal understanding of the problem being studied. Plot details such as axes, legends, color and size are cleaned up later to convey more information in an aesthetically pleasing way.

To demonstrate these ideas, we've copied some data for you from the U.S. Environmental Protection Agency (EPA) which sets national ambient air quality standards for outdoor air pollution. These Standards say that for fine particle pollution (PM2.5), the "annual mean, averaged over 3 years" my not exceed 12 micro grams per cubic meter. We stored the data from the U.S. EPA web site in the data frame `pollution`. Use the R function `head` to see the first few entries of `pollution`.

```{r a-1, exercise = TRUE}

```
```{r a-1-solution}
head(pollution)
```

You should see 6 rows of 5 columns of data. The pollution count is in the first column labeled `pm25`. We'll work mostly with that. The other 4 columns are a `fips` code indicating the state (first 2 digits) and county (last 3 digits) with that count, the associated `region` (east or west), and the `longitude` and `latitude` of the area. Now run the R command `dim` with `pollution` as an argument to see how long the table is.

```{r c, exercise = TRUE}

```
```{r c-solution}
dim(pollution)
```

So there are 576 entries in `pollution`. We'd like to investigate the question "Are there any counties in the U.S. that exceed that national standard (12 micro grams per cubic meter) for fine particle pollution?" We'll look at several one-dimensional summaries of the data to investigate this question.

The first technique uses the R command `summary`, a 5-number summary which returns 6 numbers. Run it now with the `pm25` column of `pollution` as its argument. Recall that the construct for this is `pollution$pm25`.

```{r d, exercise = TRUE}

```
```{r d-solution}
summary(pollution$pm25)
```

This shows us basic info about the `pm25` data, namely its `Minimum` (0 percentile) and `Maximum` (100 percentile) values, and three `Quartiles` of the data. These last indicate the `pollution` measures at which 25%, 50%, and 75% of the counties fall below. In addition to these 5 numbers we see the `Mean`, or average, measure of particulate pollution across the 576 counties.

```{r e, echo=FALSE}
question("Half the measured counties have a pollution level less than or equal to what number of micrograms per cubic meter?",
         answer("9.836"),
         answer("10.050", correct = TRUE),
         answer("11.360"),
         answer("8.549")
)
```

## One-Dimensional Plots

To save you a lot of typing, we've saved off `pollution$pm25` for you in the variable `ppm`. You can use `ppm` now in place of the longer expression. Try it now as the argument of the R command `quantile`. See how the results look a lot like the results of the output of the `summary` command?

```{r f, exercise = TRUE}

```
```{r f-solution}
quantile(ppm)
```

Quantile gives the quartiles, right? 

```{r g, echo=FALSE}
question("What is the one value missing from this quantile output that summary gave you?",
         answer("the mean", correct = TRUE),
         answer("the median"),
         answer("the minimum value"),
         answer("the maximum value")
)
```

### Boxplot

Now we'll plot a picture, specifically a boxplot. Run the R command `boxplot` with `ppm` as an input. Also specify the color parameter `col` equal to "blue".

```{r h, exercise = TRUE}

```
```{r h-solution}
boxplot(ppm, col = "blue")
```

The boxplot shows us the same quartile data that summary and quantile did. The lower and upper edges of the blue box respectively show the values of the 25% and 75% quantiles.

```{r i, echo=FALSE}
question("what do you think the horizontal line inside the box represents?",
         answer("the mean"),
         answer("the minimum value"),
         answer("the median", correct = TRUE),
         answer("the maximum value")
)
```

The "whiskers" of the box (the vertical lines extending above and below the box) relate to the "range" parameter of boxplot, which we let default to the value 1.5 used by R. The height of the box is the interquartile range, the difference between the 75th and 25th quantiles. In this case that difference is 2.8. The whiskers are drawn to be a length of `range*2.8` or `1.5*2.8`. This shows us roughly how many, if any, data points are outliers, that is, beyond this range of values.

Note that `boxplot` is part of R's base plotting package. A nice feature that this package provides is its ability to overlay features. That is, you can add to (annotate) an existing plot.

### Boxplot with Overlays

To see this, run the R command `abline` with the argument `h` equal to 12. Recall that 12 is the EPA standard for air quality.

```{r j, exercise = TRUE}

```
```{r j-solution}
abline(h = 12)
```
```{r k, echo=FALSE}
question("What did the command do?",
         answer("nothing"),
         answer("hid 12 random data points", correct = TRUE),
         answer("drew a vertical line at 12"),
         answer("drew a horizontal line at 12", correct = TRUE)
)
```

So `abline` "adds one or more straight lines through the current plot." We see from the plot that the bulk of the measured counties comply with the standard since they fall under the line marking that standard.

### Histograms

Now use the R command `hist` (another function from the base package) with the argument `ppm`. Specify the color parameter `col` equal to "green". This will plot a histogram of the data.

```{r l, exercise = TRUE}

```
```{r l-solution}
hist(ppm, col = "green")
```

The histogram gives us a little more detailed information about our data, specifically the distribution of the pollution counts, or how many counties fall into each bucket of measurements.

```{r m, echo=FALSE}
question("What are teh most frequent pollution counts?",
         answer("under 5"),
         answer("between 12 and 14"),
         answer("between 9 and 12", correct = TRUE),
         answer("between 6 and 8")
)
```

### Histograms with rugs

Now run the R command `rug` with the argument `ppm`.

```{r n, exercise = TRUE}

```
```{r n-solution}
rug(ppm)
```

This one-dimensional plot, with its grayscale representation, gives you a little more detailed information about how many data points are in each bucket and where they lie within the bucket. It shows (through density of tick marks) that the greatest concentration of counties has between 9 and 12 micrograms per cubic meter just as the histogram did.

To illustrate this a little more, we've defined for you two vectors, `high` and `low`, containing `pollution` data of high (greater than 15) and low (less than 5) values respectively. Look at `low` now and see how it relates to the output of `rug`.

```{r o, exercise = TRUE}

```
```{r o-solution}
low
```

It confirms that there are two data points between 3 and 4 and many between 4 and 5. Now look at high.

```{r p, exercise = TRUE}

```
```{r p-solution}
high
```

Again, we see one data point greater than 18, one between 17 and 18, several between 16 and 17 and two between 15 and 16, verifying what `rug` indicated.

Now rerun `hist` with 3 arguments, `ppm` as its first, `col` equal to "green", and the argument `breaks` equal to 100.

```{r q, exercise = TRUE}

```
```{r q-solution}
hist(ppm, col = "green", breaks = 100)
```
```{r r, echo=FALSE}
question("What do you think the breaks argument specifies in this case?",
         answer("the number of buckets to split the data into", correct = TRUE),
         answer("the number of data points to graph"),
         answer("the number of stars in the sky"),
         answer("the number of counties exceeding the EPA standard")
)
```

So this histogram with more buckets is not nearly as smooth as the preceding one. In fact, it's a little too noisy to see the distribution clearly. When you're plotting histograms you might have to experiment with the argument `breaks` to get a good idea of your data's distribution. For fun now, rerun the R command `rug` with the argument `ppm`.

```{r s, exercise = TRUE}

```
```{r s-solution}
rug(ppm)
```

See how rug works with the existing plot? It automatically adjusted its pocket size to that of the last plot plotted.

Now rerun `hist` with `ppm` as the data and `col` equal to "green".

```{r t, exercise = TRUE}

```
```{r t-solution}
hist(ppm, col = "green")
```

### Histograms with rugs and Overlays

Now run the command `abline` with the argument `v` equal to 12 and the argument `lwd` equal to 2.

```{r u, exercise = TRUE}

```
```{r u-solution}
abline(v = 12, lwd = 2)
```

See the vertical line at 12? Not very visible, is it, even though you specified a line width of 2? Run `abline` with the argument `v` equal to `median(ppm)`, the argument `col` equal to "magenta", and the argument `lwd` equal to 4.

```{r v, exercise = TRUE}

```
```{r v-solution}
abline(v = median(ppm), col = "magenta", lwd = 4)
```

Better, right? Thicker and more of a contrast in color. This shows that although the median (50%) is below the standard, there are a fair number of counties in the U.S that have pollution levels higher than the recommended standard.

Now recall that our `pollution` data had 5 columns of information. So far we've looked only at the `pm25` column. We can also look at other information. To remind yourself what's there run the R command `names` with `pollution` as the argument.

```{r w, exercise = TRUE}

```
```{r w-solution}
names(pollution)
```

`longitude` and `latitude` don't sound interesting, and each `fips` is unique since it identifies states (first 2 digits) and counties (last 3 digits). Let's look at the `region` column to see what's there. Run the R command `table` on this column. Use the construct `pollution$region`. Store the result in the variable `reg`.

### Barplots

```{r x, exercise = TRUE}

```
```{r x-solution}
reg <- table(pollution$region)
```

Look at `reg`.

```{r prepare-y}
reg <- table(pollution$region)
```
```{r y, exercise=TRUE}

```
```{r y-solution}
reg
```

Lot more counties in the east than west. We'll use the R command `barplot` (another type of one-dimensional summary) to plot this information. Call `barplot` with `reg` as its first argument, the argument `col` equal to "wheat", and the argument `main` equal to the string "Number of Counties in Each Region".

```{r z, exercise = TRUE, exercise.setup = "prepare-y"}

```
```{r z-solution}
barplot(reg, col = "wheat", main = "Number of Counties in Each Region")
```

```{r aa, echo=FALSE}
question("What do you think the argument main specifies?",
         answer("the y axis label"),
         answer("the title of the graph", correct = TRUE),
         answer("the x axis label"),
         answer("I can't tell")
)
```

So we've seen several examples of one-dimensional graphs that summarize data. Two dimensional graphs include scatterplots, multiple graphs which we'll see more examples of, and overlayed one-dimensional plots which the R packages such as `lattice` and `ggplot2` provide.

## Multi-Dimensional Plots

Some graphs have more than two dimensions. These include overlayed or multiple two-dimensional plots and spinning plots. Some three-dimensional plots are tricky to understand so have limited applications. We'll see some examples now of more complicated graphs, in particular, we'll show two graphs together.

### Multi-Dimensional Boxplots

First we'll show how R, in one line and using base plotting, can display multiple boxplots. We simply specify that we want to see the `pollution` data as a function of `region`. We know that our `pollution` data characterized each of the 576 entries as belonging to one of two regions (east and west).

We use the R formula `y ~ x` to show that `y` (in this case `pm25`) depends on `x` (`region`). Since both come from the same data frame (`pollution`) we can specify a data argument set equal to `pollution`. By doing this, we don't have to type `pollution$pm25` (or `ppm`) and `pollution$region`. We can just specify the formula `pm25 ~ region`. Call `boxplot` now with this formula as its argument, data equal to `pollution`, and `col` equal to "red".

```{r ab, exercise = TRUE}

```
```{r ab-solution}
boxplot(pm25 ~ region, data = pollution, col = "red")
```

Two for the price of one! 

### Multi-Dimensional Histograms

Similarly, we can plot multiple histograms in one plot, though to do this we have to use more than one R command. First we have to set up the plot window with the R command `par`, which specifies how we want to lay out the plots---say, one above the other. We also use `par` to specify the margins, passed in a 4-long vector which indicates the number of lines for the bottom, left, top and right. Type the R command `par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))` now. Don't expect to see any new result.

```{r ac, exercise = TRUE}

```
```{r ac-solution}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
```

So we set up the plot window for two rows and one column with the `mfrow` argument. The `mar` argument set up the margins. Before we plot the histograms let's explore the R command `subset` which, not surprisingly, "returns subsets of vectors, matrices or data frames which meet conditions". We'll use `subset` to pull off the data we want to plot. Call `subset` now with `pollution` as its first argument, followed by a boolean expression that tests whether `region` equals the string "east". Put the result in the variable `east`.

```{r prepare-ac-1}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
```
```{r ac-1, exercise=TRUE}

```
```{r ac-1-solution}
east <- subset(pollution, region == "east")
```

Use `head` to look at the first few entries of east.

```{r prepare-ae}
east <- subset(pollution, region == "east")
west <- subset(pollution, region == "west")
```
```{r ae, exercise=TRUE}

```
```{r ae-solution}
head(east)
```

So `east` holds more information than we need. We just want to plot a histogram with the `pm25` portion. Call `hist` now with the `pm25` portion of `east` as its first argument and `col` equal to "green" as its second.

```{r ae-1, exercise = TRUE, exercise.setup = "prepare-ac-1"}

```
```{r ae-1-solution}
hist(east$pm25, col = "green")
```

See? The command `par` told R we were going to have one column with 2 rows, so it placed this histogram in the top position.

Now, here's a challenge for you. Plot the histogram of the counties from the west using just one R command. Let the appropriate `subset` command (with the `pm25` portion specified) be the first argument and `col` (equal to "green") the second.  Replace "east" with the `subset` command. Make sure the boolean argument checks for equality between `region` and "west".

```{r ag, exercise = TRUE, exercise.setup = "prepare-ae"}

```
```{r ag-solution}
hist(subset(pollution, region == "west")$pm25, col = "green")
```

See how R does all the labeling for you? Notice that the titles are different since we used different commands for the two plots. Let's look at some scatter plots now.

### Scatter Plots (Which are, by Nature, Multi-Dimensional [Most of the Time])

Scatter plots are two-dimensional plots which show the relationship between two variables, usually `x` and `y`. Let's look at a scatterplot showing the relationship between `latitude` and the `pm25` data. We'll use `plot`, a function from R's base plotting package.

We've seen that we can use a function call as an argument when calling another function. We'll do this again when we call `plot` with the arguments `latitude` and `pm25`, which are both from our data frame `pollution`. We'll call `plot` from inside the R command `with` which evaluates "an R expression in an environment constructed from data". We'll use `pollution` as the first argument to `with` and the call to `plot` as the second. This allows us to avoid typing "`pollution$`" before the arguments to plot, so it saves us some typing and adds to your base of R knowledge. Try this now.

```{r ah, exercise = TRUE}

```
```{r ah-solution}
with(pollution, plot(latitude, pm25))
```

Note that the first argument is plotted along the x-axis and the second along the y. Now use `abline` to add a horizontal line at 12. Use two additional arguments, `lwd` equal to 2 and `lty` also equal to 2. See what happens.

```{r prepare-ai}
with(pollution, plot(latitude, pm25))
```
```{r ai, exercise=TRUE}

```
```{r ai-solution}
abline(h = 12, lwd = 2, lty = 2)
```

See how `lty = 2` made the line dashed? Now let's replot the scatterplot. This time, instead of using `with`, call `plot` directly with 3 arguments. The first 2 are `pollution$latitude` and `ppm`. The third argument, `col`, we'll use to add color and more information to our plot. Set this argument (`col`) equal to `pollution$region` and see what happens.

```{r aj, exercise=TRUE}

```
```{r aj-solution}
plot(pollution$latitude, ppm, col = pollution$region)
```

We've got two colors on the map to distinguish between counties in the east and those in the west. Can we figure out which color is `east` and which `west`? See that the high (greater than 50) and low (less than 25) latitudes are both red.

```{r ak, echo=FALSE}
question("Latitudes indicate distance from the equator, so which half of the U.S. (east or west) has counties at the extreme north and south?",
         answer("east"),
         answer("west", correct = TRUE)
)
```

As before, use `abline` to add a horizontal line at 12. Use two additional arguments, `lwd` equal to 2 and `lty` also equal to 2.

```{r prepare-al}
plot(pollution$latitude, ppm, col = pollution$region)
```
```{r al, exercise=TRUE}

```
```{r al-solution}
abline(h = 12, lwd = 2, lty = 2)
```

We see many counties are above the healthy standard set by the EPA, but it's hard to tell overall, which `region`, `east` or `west`, is worse.

Let's plot two scatterplots distinguished by `region`.

As we did with multiple histograms, we first have to set up the plot window with the R command `par`. This time, let's plot the scatterplots side by side (one row and two columns). We also need to use different margins. Type the R command `par(mfrow = c(1, 2), mar = c(5, 4, 2, 1))` now. Don't expect to see any new result.

```{r am, exercise = TRUE}

```
```{r am-solution}
par(mfrow = c(1, 2), mar = c(5, 4, 2, 1))
```

For the first scatterplot, on the left, we'll plot the `latitude`s and `pm25` counts from the west. We already pulled out the information for the counties in the east. Let's now get the information for the counties from the west. Create the variable `west` by using the `subset` command with `pollution` as the first argument and the appropriate boolean as the second.

```{r an, exercise = TRUE}

```
```{r an-solution}
west <- subset(pollution, region == "west")
```

Now call `plot` with three arguments. These are `west$latitude`, `west$pm25` (y-axis), and the argument `main` equal to the string "West" (title). Do this now.

```{r prepare-ao}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
east <- subset(pollution, region == "east")
west <- subset(pollution, region == "west")
```
```{r ao, exercise=TRUE}

```
```{r ao-solution}
plot(west$latitude, west$pm25, main = "West")
```

For the second scatterplot, on the right, we'll plot the latitudes and pm25 counts from the east.

```{r prepare-ap}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
east <- subset(pollution, region == "east")
west <- subset(pollution, region == "west")
plot(west$latitude, west$pm25, main = "West")
```
```{r ap, exercise=TRUE}

```
```{r ap-solution}
plot(east$latitude, east$pm25, main = "East")
```

See how R took care of all the details for you? Nice, right? It looks like there are more dirty counties in the east but the extreme dirt (greater than 15) is in the west.

## Summary and Review

Let's summarize and review.

```{r aq, echo=FALSE}
question("Which of the following characterizes exploratory plots?
",
         answer("quick and dead"),
         answer("slow and steady"),
         answer("quick and dirty", correct = TRUE),
         answer("slow and clean")
)
```
```{r ar, echo=FALSE}
question("Plots let you summarize the data (usually graphically) and highlight any broad features",
         answer("True", correct = TRUE),
         answer("False")
)
```
```{r as, echo=FALSE}
question("Which of the following do plots NOT do?",
         answer("Summarize the data (usually graphically) and highlight any broad features"),
         answer("Conclude that you are ALWAYS right", correct = TRUE),
         answer("Suggest modeling strategies for the 'next step'"),
         answer("Explore basic questions and hypotheses (and perhaps rule them out)")
)
```

Congrats! You've concluded exploring this lesson on graphics. We hope you didn't find it too quick or dirty.